---
title: "Confirmatory Data Analysis"
author: "Xing Song"
date: "10/22/2020"
output: html_document
---

Preparation: report setup, load packages, load data

```{r setup}
#set up rmd 
knitr::opts_chunk$set(message=F,warning=F,fig.width=6, fig.height=3)
pacman::p_load(tidyverse,epitools,rcompanion)
idd_bmi<-readRDS("./idd_bmi_final.rda")
```


**************************************************************************************************

## Confirmatory Data Analysis (CDA) uses inferential statistics to:

- Evaluate evidence by challenging their assumptions about the data.
- Produce estimates with a specified level of precision
- Test Hypotheses about correlations
- Estimate relationships between variables
- Evaluate the difference between 


**************************************************************************************************

```{r}
##==============================================
# "str" function is a convenient function to 
# quickly check data structure and data types
##==============================================
str(idd_bmi)
```


For easy demonstration, I added the following three categorical variables: `bmi_class`, `obesity_ind`, `down_ind`

```{r bmi_classify}
idd_bmi<-idd_bmi %>%
  filter(DIAGNOSIS!="both") %>%
  mutate(bmi_class=case_when(bmi_median<18.5 ~ '1.underweight',
                             bmi_median>=18.5&bmi_median < 25 ~ "2.normal",
                             bmi_median>=25&bmi_median<30 ~ "3.overweight",
                             bmi_median>=30 ~ "4.obese"),
         obesity_ind=case_when(bmi_median>=30 ~ 1,
                               TRUE ~ 0),
         down_ind=case_when(DIAGNOSIS=="down" ~ 1,
                            TRUE ~ 0))
```

*Motivation Research Question: I want to examine if IDD patients with down syndrom is more likely to be obese (or with higher BMI) than patients with autism.*


#### Categorical Exposure and Categorical Outcome

```{r}
ds_obese_tbl<-table(idd_bmi$down_ind,idd_bmi$obesity_ind,dnn=list("With DS","Is_obese"))
ds_obese_tbl2<-as.data.frame(addmargins(ds_obese_tbl))
```

*Odds Ratio*

Odds Ratio calculation in details: 

```{r oddsratio}
ds1_obese1<-ds_obese_tbl2$Freq[5]
ds0_obese1<-ds_obese_tbl2$Freq[4]

ds1_obese0<-ds_obese_tbl2$Freq[2]
ds0_obese0<-ds_obese_tbl2$Freq[1]

odds_obese1<-ds1_obese1/ds0_obese1
odds_obese0<-ds1_obese0/ds0_obese0

odds_ratio<-odds_obese1/odds_obese0
odds_ratio 
```

```{r}
epitools::oddsratio(ds_obese_tbl)
```


*Phi Correlation Coefficients*
```{r}
rcompanion::phi(ds_obese_tbl)
```

*Cramer's V statistics*

```{r cramerv}
rcompanion::cramerV(ds_obese_tbl)
```

************************************************************************************************

#### Categorical Exposure and Numerical Outcome

*t-statistics*

```{r t}
t.test(bmi_median ~ DIAGNOSIS, data=idd_bmi)
```

*ANOVA*

```{r anova}
summary(aov(bmi_median ~ DIAGNOSIS, data=idd_bmi))
```

*Kruskal-Wallis Test*

```{r kruskal}
kruskal.test(bmi_median ~ DIAGNOSIS, data=idd_bmi)
```
*******************************************************************************************************

#### Numerical Exposure and Numerical Outcome

*Pearson Correlation Coefficient*

```{r pearson}
cor.test(idd_bmi$bmi_median,idd_bmi$AGE_AT_FIRSTDX,method="pearson")
```


*Spearman Correlation Coefficient*

```{r spearman}
cor.test(idd_bmi$bmi_median,idd_bmi$AGE_AT_FIRSTDX,method="spearman")
```


