---
title: "Introduction to Unsupervised Learning"
author: "Xing Song"
date: "11/11/2020"
output: html_document
---

Let's do some preparation for the class by running the following chunk of codes

```{r setup}
#set up rmd 
knitr::opts_chunk$set(message=F,
                      warning=F,
                      echo=T,
                      fig.width=6,
                      fig.height=4)

#load packages
pacman::p_load(tidyverse,
               factoextra #for cluster, pca visualization
               )
```


```{r}
#load sample data
idd_bmi<-readRDS("./idd_bmi_final.rda") 
str(idd_bmi)

#let's only take numerical variables
idd_bmi_num<-idd_bmi %>%
  select(AGE_AT_FIRSTDX,bmi_median,OSA_ind,
         sex_f_ind,`amerian ind`,asian,black,other,`two races`,unknown,white)
colnames(idd_bmi_num)<-gsub(" ","_",colnames(idd_bmi_num))
```


## K-Means Clustering 

`kmeans` function parameters: 

- `x`: numeric matrix, numeric data frame or a numeric vector
- `centers`: Possible values are the number of clusters (k) or a set of initial (distinct) cluster centers. If a number, a random set of (distinct) rows in x is chosen as the initial centers.
- `iter.max`: The maximum number of iterations allowed. Default value is 10.
- `nstart`: The number of random starting partitions when centers is a number. Trying nstart > 1 is often recommended.

```{r}
#scale the data
idd_bmi_scale<-scale(idd_bmi_num,center=T,scale=T) #z-score scaling

#try two clusters
set.seed(123)
km2<-kmeans(x=idd_bmi_scale,
            center=2,
            iter.max = 15,
            nstart = 10)

#visualize
fviz_cluster(km2, 
             data = idd_bmi_scale,
             palette = c("#2E9FDF", "#00AFBB"), 
             geom = "point",
             ellipse.type = "convex", 
             ggtheme = theme_bw()
             )
```

### Principal Component Analysis (PCA)

The idea of PCA is simple — reduce the number of variables of a data set, while preserving as much information as possible

- Each principal component is a linear combination of original variables.
- PCA is useful for visualization, because it can plot high-dimensional data along first two directions of maximum variance.
- Data should be centered and scaled to unit-variance

```{r}
# Dimension reduction using PCA
pc <- prcomp(~ AGE_AT_FIRSTDX+bmi_median+OSA_ind+sex_f_ind+`amerian_ind`+asian+black+other+`two_races`+unknown+white, data = idd_bmi_num, scale = TRUE, center=TRUE)

summary(pc)

fviz_eig(pc)
```


```{r, fig.width=8,fig.height=6}
fviz_pca_var(pc,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```


```{r}
# Coordinates of individuals
ind.coord <- as.data.frame(get_pca_ind(pc)$coord)

ggplot(ind.coord,aes(x=Dim.1,y=Dim.2))+
  geom_point()
```


### Estimating the optimal number of clusters
> One fundamental question is: How to choose the right number of expected clusters (k)?

##### Elbow method
Distance metric based on the `inertia` or `within sum of square`, which is a measure of average distance within a cluster. 

```{r}
fviz_nbclust(idd_bmi_scale, kmeans, method = "wss") +
    geom_vline(xintercept = 4, linetype = 2)+
  labs(subtitle = "Elbow method")
```


##### Silhouette method
Distance metric based on the `silhouette value`, which is a measure of how similar an object is to its own cluster (cohesion) compared to other clusters (separation)

```{r}
fviz_nbclust(idd_bmi_scale, kmeans, method = "silhouette")+
  labs(subtitle = "Silhouette method")
```


##### Gap statistic
The `Gap statistic standardizes` the graph of log(wss), where wss is the within-cluster dispersion, by comparing it to its expectation under an appropriate null reference distribution of the data

```{r}
# nboot = 50 to keep the function speedy. 
# recommended value: nboot= 500 for your analysis (may take more than 1hr).
# Use verbose = FALSE to hide computing progression.

set.seed(123)
fviz_nbclust(idd_bmi_scale, kmeans, nstart = 10,  method = "gap_stat", nboot = 20)+
  labs(subtitle = "Gap statistic method")
```

Remark: The disadvantage of elbow and average silhouette methods is that, they measure a global clustering characteristic only. A more sophisticated method is to use the gap statistic which provides a statistical procedure to formalize the elbow/silhouette heuristic in order to estimate the optimal number of clusters.

```{r}
# Add clusters obtained using the K-means algorithm
ind.coord$cluster <- factor(km2$cluster)

ggplot(ind.coord,aes(x=Dim.1,y=Dim.2))+
  geom_point(aes(color=cluster))
```


## Hierarchical Agglomeration Clustering

#### Similarity and Distance Metrics

*Euclidean Distance*

$$ Euclidean(p_1,p_2)=\sqrt((x_1-x_2)^2+(y_1-y_2)^2  $$

```{r, fig.width=12,fig.height=8}
hc_idd <- idd_bmi_num   %>%  # Get cars data
          dist(method="euclidean")   %>%  # Compute distance/dissimilarity matrix
          hclust      # Computer hierarchical clusters
  
plot(hc_idd)          # Plot dendrogram
rect.hclust(hc_idd, k = 2, border = "red")
rect.hclust(hc_idd, k = 3, border = "blue")
```


*Manhattan Distance*

$$ Manhattan(p_1,p_2)=|x_1-x_2|+|y_1-y_2|  $$

```{r, fig.width=12,fig.height=8}
hc_idd <- idd_bmi_num   %>%  # Get cars data
          dist(method="manhattan")   %>%  # Compute distance/dissimilarity matrix
          hclust      # Computer hierarchical clusters
  
plot(hc_idd)          # Plot dendrogram
rect.hclust(hc_idd, k = 2, border = "red")
rect.hclust(hc_idd, k = 3, border = "blue")
```


Other more advanced methods: 

- t-Distributed Stochastic Neighbor Embedding (t-SNE)
- Mixture Models
- Multidimensional Scaling (MDS)
- Non-negative Matrix Factorization (NMF)
- Auto-encoders (Deep Learning)



